\subsection{Applying PCA}

What can we do with PCA?

\begin{enumerate}
\item Dimensionality reduction
\item Visualization (e.g. plot the projections onto the first 2 or 3 PCs.)
\item Whitening (i.e. decorrelate the data) - exploit that eigenvectors form an orthonormal basis.
$$
\vec v^{(\alpha)} = \vec x_{whitened}^{(\alpha)} = \vec{\Lambda}^{-\frac{1}{2}}\vec{M}^\top\vec{x}^{(\alpha)}
\quad
\forall \alpha
$$

\question{What is the covariance of $\vec x_{whitened}$?}

\end{enumerate}

\newpage
\subsection{A note on implementation:}

\question{Is an off-the-shelf implementation such as\\ sklearn.decomposition.PCA for python really necessary?}

PCA boils down to solving the eigenvalue problem.
We bascially need a routine with which to solve the eigenvalue problem. For python, the obvious choice is to use a routine like \textit{numpy.linalg.eig(A)}. \emph{Whatch out for how to intepret the output.}

\question{Are there alternatives to eigenvalue decomposition?}

Yes, namely,

\underline{Singular Value Decomposition (SVD)}:\\
Let $\vec X' \in \R^{N \times p}$ be the centered dataset (centering is a must):
$$
\vec X'_{N \times p} = \vec U_{N \times N} \, \vec S_{N \times p} \, \vec V^\top_{{p \times p}}
$$
where,
$$
\vec U^\top \vec U = \vec I_{N \times N}
$$ and 
$$
\vec V^\top \vec V = \vec I_{p \times p}
$$ (i,e. $\vec U$, $\vec V$ are orthogonal.).
\begin{itemize}
\item The columns of $\vec U$ make up the eigenvectors of $\vec X'\vec X'^\top$.
\item The columns of $\vec V$ (or rows of $\vec V^\top$) make up the eigenvectors of $\vec X'^\top\vec X'$.
\item $\vec S$ is a diagonal matrix. The singular values in $\vec S$ (the diagonal entries) are the \textbf{square roots} of the  eigenvalues of the eigenvectors of $\vec X'\vec X'^\top$ or $\vec X'^\top\vec X'$.

\end{itemize}
Recall:

$$
\vec C = \E[\vec X \vec X^\top] = \frac{1}{p} \vec X \vec X^\top
$$

\underline{When to use SVD vs. Eig? \textbf{Not dogma}}

\begin{itemize}

\item Numerical instabilities in eigendecomposition of very large covariance matrix $\rightarrow$ SVD
\item SVD not applicable to Kernel-PCA $\rightarrow$ Eig
\item Computational efficiency $\rightarrow$ SVD (depends...?)
\end{itemize}
